---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
googlesheets4::sheets_deauth()
```

# googlesheets4

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/googlesheets4)](https://CRAN.R-project.org/package=googlesheets4)
[![Travis build status](https://travis-ci.org/tidyverse/googlesheets4.svg?branch=master)](https://travis-ci.org/tidyverse/googlesheets4)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/tidyverse/googlesheets4?branch=master&svg=true)](https://ci.appveyor.com/project/tidyverse/googlesheets4)
[![Coverage status](https://codecov.io/gh/tidyverse/googlesheets4/branch/master/graph/badge.svg)](https://codecov.io/github/tidyverse/googlesheets4?branch=master)
<!-- badges: end -->

googlesheets4 provides an R interface to [Google Sheets](https://spreadsheets.google.com/) via the [Sheets API v4](https://developers.google.com/sheets/api/). It is a reboot of the existing [googlesheets package](https://cran.r-project.org/package=googlesheets).

*Why **4**? Why googlesheets**4**? Did I miss googlesheets1 through 3? No. The idea is to name the package after the corresponding version of the Sheets API. In hindsight, the original googlesheets should have been googlesheets**3**.*

The best source of information is always the package website: [googlesheets4.tidyverse.org](https://googlesheets4.tidyverse.org) 

## Installation

You can install the released version of googlesheets4 from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("googlesheets4")
```

And the development version from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("tidyverse/googlesheets4")
```

## Load googlesheets4

```{r}
library(googlesheets4)
```

## Auth

googlesheets4 will, by default, help you interact with Sheets as an authenticated Google user. The package facilitates this process upon first need. For this overview, we've logged into Google as a specific user in a hidden chunk. For more about auth, visit the package website: [googlesheets4.tidyverse.org](https://googlesheets4.tidyverse.org).

## `read_sheet()`

`read_sheet()` is the main "read" function and should evoke `readr::read_csv()` and `readxl::read_excel()`. It's an alias for `sheets_read()`. Most functions in googlesheets4 actually start with `sheets_`. googlesheets4 is pipe-friendly (and reexports `%>%`), but works just fine without the pipe.

We demonstrate basic functionality using some world-readable example sheets accessed via `sheets_examples()` and `sheets_example()`.

Read everything:

```{r}
sheets_example("chicken-sheet") %>% 
  read_sheet() # or use sheets_read()
```

Read specific cells, from a specific sheet, using an A1-style notation:
```{r}
sheets_example("deaths") %>% 
  read_sheet(range = "arts!A5:F15")
```

Read from a named range or region and specify (some of the ) column types:
```{r}
sheets_example("deaths") %>% 
  read_sheet(range = "arts_data", col_types = "??i?DD")
```

There are various ways to specify the target Sheet. The simplest, but ugliest, is to provide the URL.

```{r}
# url of the 'chicken-sheet' example
url <- "https://docs.google.com/spreadsheets/d/1ct9t1Efv8pAGN9YO5gC2QfRq2wT4XjNoTMXpVeUghJU"
read_sheet(url)
```

For more information, see the package website: [googlesheets4.tidyverse.org](https://googlesheets4.tidyverse.org).

## What's yet to come?

Writing into Sheets. The googledrive package 
([googledrive.tidyverse.org](https://googledrive.tidyverse.org)) can already be used to write into Sheets at the "whole file" level, because that is carried out via the Drive API. `googledrive::drive_upload()` and `googledrive::drive_update()` are very useful for this.

But, if you need more granular control, such as writing to specific worksheets or cells, that requires the Sheets API. This is not yet implemented in googlesheets4, but will be.

## Contributing

If you'd like to contribute to the development of googlesheets4, please read [these guidelines](https://googlesheets4.tidyverse.org/CONTRIBUTING.html).

Please note that the 'googlesheets4' project is released with a [Contributor Code of Conduct](https://googlesheets4.tidyverse.org/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Privacy

[Privacy policy](https://www.tidyverse.org/google_privacy_policy)

## Context

googlesheets4 draws on and complements / emulates other packages in the tidyverse:

  * [googlesheets](https://cran.r-project.org/package=googlesheets) is the package that googlesheets4 is replacing. Main improvements in googlesheets4: (1) wraps the current, most modern Sheets API; (2) leans on googledrive for all "whole file" operations; and (3) uses shared infrastructure for auth and more, from the gargle package. Main deficiency: googlesheets4 doesn't _write_ yet.
  * [googledrive](http://googledrive.tidyverse.org) already provides a fully-featured interface to the Google Drive API. Any "whole file" operations can already be accomplished *today* with googledrive: upload or download or update a spreadsheet, copy, rename, move, change permission, delete, etc. googledrive already supports Team Drives.
  * [readxl](http://readxl.tidyverse.org) is the tidyverse package for reading Excel files (xls or xlsx) into an R data frame. googlesheets4 takes cues from parts of the readxl interface, especially around specifying which cells to read.
  * [readr](http://readr.tidyverse.org) is the tidyverse package for reading delimited files (e.g., csv or tsv) into an R data frame. googlesheets4 takes cues from readr with respect to column type specification.
