---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/tidyverse/googlesheets4.svg?branch=master)](https://travis-ci.org/tidyverse/googlesheets4)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/tidyverse/googlesheets4?branch=master&svg=true)](https://ci.appveyor.com/project/tidyverse/googlesheets4)
[![Coverage status](https://codecov.io/gh/tidyverse/googlesheets4/branch/master/graph/badge.svg)](https://codecov.io/github/tidyverse/googlesheets4?branch=master)

# googlesheets4

googlesheets4 provides an R interface to [Google Sheets](https://spreadsheets.google.com/) via the [Sheets API v4](https://developers.google.com/sheets/api/). It is a reboot of the existing [googlesheets package](https://cran.r-project.org/package=googlesheets).

*Why **4**? Why googlesheets**4**? Did I miss googlesheets1 through 3? No. The idea is to name the package after the corresponding version of the Sheets API. In hindsight, the original googlesheets should have been googlesheets**3**.*

## Installation

You can install the released version of googlesheets4 from [CRAN](https://CRAN.R-project.org) with:

``` r
## NO, NO YOU CANNOT
## install.packages("googlesheets4")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tidyverse/googlesheets4")
```

## No auth yet!

Sorry, auth hasn't been wired up yet, but that's the next priority. Until then, you can only use googlesheets4 to access Sheets where sharing settings say: "anyone with a link can view". `read_sheet()` has an example that shows googledrive calls to achieve this or you can do in the Sheets browser UI via *File > Share ...*.

## Read a Sheet

`read_sheet()` is the main "read" function and should evoke `readr::read_csv()` and `readxl::read_excel()`. It's an alias for `sheets_read()` (most functions in googlesheets4 actually start with `sheets_`). googlesheets4 is pipe-friendly (and rexports `%>%`), but works just fine without the pipe.

### Identify and read your own Sheet

Let's say you have a cheerful Google Sheet named "deaths". If you want to access it by name, use googledrive to identify the document (capture its metadata, espcially file id). Pass the result to functions like `sheets_get()` (gets file metadata) or `read_sheet()` (gets cell data).

```{r}
library(googledrive)
library(googlesheets4)

(deaths <- drive_get("deaths"))
sheets_get(deaths)
```

If you're willing to deal with the sheet's id, just provide that directly to googlesheets4 functions and omit googledrive.

```{r}
sheets_get("1ESTf_tH08qzWwFYRC1NVWJjswtLdZn9EGw5e3Z5wMzA")
```

Lesson: googledrive is the friendliest way to work with files on Google Drive, including those that are Google Sheets. You can refer to files by name. googlesheets4 is focused on operrations specific to Sheets and is more programming oriented. You must pass a file id or something that contains the file id.

### Specifying which cells to read

We've made a few Sheets easy to access via `sheets_example()`. Here we read from a mini-Gapminder Sheet to show some of the different ways to specify (work)sheet and cell ranges. Note also that `col_types` gives control of column types.

```{r}
library(googlesheets4)

read_sheet(sheets_example("mini-gap"), sheet = 2)

read_sheet(sheets_example("mini-gap"), sheet = "Oceania", n_max = 3)

read_sheet(sheets_example("deaths"), skip = 4, n_max = 10)

read_sheet(
  sheets_example("deaths"), range = "other!A5:F15", col_types = "?ci??D"
)
```

## Other functions

googlesheets4 exposes Sheet metadata via `sheets_get()` and can also be used to access raw cell data (one row per cell) via `sheets_cells()`. That data can be post-processed with `spread_sheet()` to obtain the same data frame you get from `read_sheet()`.

```{r}
sheets_get(sheets_example("mini-gap"))

(df <- sheets_cells(sheets_example("deaths"), range = "E5:E7"))
df$cell[[3]]

df %>% spread_sheet(col_types = "D")
## is same as ...
read_sheet(sheets_example("deaths"), range = "E5:E7", col_types ="D")
```
  
## What's coming soon?

OAuth2

Writing to Sheets

## Context

googlesheets4 draws on and complements / emulates other packages in the tidyverse:

  * [googledrive](http://googledrive.tidyverse.org) already provides a fully-featured interface to the Google Drive API. Any "whole file" operations can already be accomplished *today* with googledrive: upload or download or update a spreadsheet, copy, rename, move, change permission, delete, etc. googledrive already supports OAuth2 and Team Drives.
  * [readxl](http://readxl.tidyverse.org) is the tidyverse package for reading Excel files (xls or xlsx) into an R data frame. googlesheets4 takes cues from parts of the readxl interface, especially around specifying which cells to read.
  * [readr](http://readr.tidyverse.org) is the tidyverse package fro reading delimited files (e.g., csv or tsv) into an R data frame. googlesheets4 takes cues from readr with respect to column type specification.

*Please note that this project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).
By participating in this project you agree to abide by its terms.*
