---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
googledrive::drive_auth(email = "jenny@rstudio.com")
googlesheets4::sheets_auth(email = "jenny@rstudio.com")
```

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/tidyverse/googlesheets4.svg?branch=master)](https://travis-ci.org/tidyverse/googlesheets4)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/tidyverse/googlesheets4?branch=master&svg=true)](https://ci.appveyor.com/project/tidyverse/googlesheets4)
[![Coverage status](https://codecov.io/gh/tidyverse/googlesheets4/branch/master/graph/badge.svg)](https://codecov.io/github/tidyverse/googlesheets4?branch=master)

# googlesheets4

googlesheets4 provides an R interface to [Google Sheets](https://spreadsheets.google.com/) via the [Sheets API v4](https://developers.google.com/sheets/api/). It is a reboot of the existing [googlesheets package](https://cran.r-project.org/package=googlesheets).

*Why **4**? Why googlesheets**4**? Did I miss googlesheets1 through 3? No. The idea is to name the package after the corresponding version of the Sheets API. In hindsight, the original googlesheets should have been googlesheets**3**.*

## Installation

You can install the released version of googlesheets4 from [CRAN](https://CRAN.R-project.org) with:

``` r
## NO, NO YOU CANNOT
## install.packages("googlesheets4")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tidyverse/googlesheets4")
```
## Auth

googlesheets4 will, by default, help you interact with Sheets as an authenticated Google user. The package facilitates this process upon first need.

The `sheets_auth_*()` family of functions gives the user more control, in order to build more advanced workflows.

For this README, we've logged into Google as a specific user in a hidden chunk. This OAuth flow is still under development and is expected to change.

## `read_sheet()`

`read_sheet()` is the main "read" function and should evoke `readr::read_csv()` and `readxl::read_excel()`. It's an alias for `sheets_read()`. Most functions in googlesheets4 actually start with `sheets_`. googlesheets4 is pipe-friendly (and reexports `%>%`), but works just fine without the pipe.

### Identify and access your own Sheet

Let's say you have a cheerful Google Sheet named "deaths". If you want to access it by name, use [googledrive](https://googledrive.tidyverse.org) to identify the document (capture its metadata, especially file id).

<!-- remove the 'message = 4' later -->

```{r message=4}
library(googledrive)
library(googlesheets4)

(deaths <- drive_get("deaths"))
```

Pass the result to googlesheets4 functions such as:

  * `sheets_get()`: gets spreadsheet-specific metadata
  * `read_sheet()`: reads cells into a data frame

```{r}
sheets_get(deaths)

read_sheet(deaths, range = "A5:F8")
```

If you're willing to deal with the spreadsheet's id, just provide that directly to googlesheets4 functions and omit googledrive.

```{r}
sheets_get("1ESTf_tH08qzWwFYRC1NVWJjswtLdZn9EGw5e3Z5wMzA")
```

Lesson: googledrive is the friendliest way to work with files on Google Drive, including files that are Google Sheets. You can refer to files by name. googlesheets4 is focused on operations specific to Sheets and is more programming oriented. You must pass a file id or something that contains the file id.

### Specify the range and column types

We've made a few world-readable Sheets easy to access via `sheets_example()`. Here we read from a mini-Gapminder Sheet to show some of the different ways to specify (work)sheet and cell ranges. Note also that `col_types` gives control of column types.

```{r}
library(googlesheets4)

read_sheet(sheets_example("mini-gap"), sheet = 2)

read_sheet(sheets_example("mini-gap"), sheet = "Oceania", n_max = 3)

read_sheet(sheets_example("deaths"), skip = 4, n_max = 10)

read_sheet(
  sheets_example("deaths"), range = "other!A5:F15", col_types = "?ci??D"
)
```

## Roundtripping with a private Sheet

Here is a demo of putting the iris data into a new, private Sheet. Then reading it back into R and exporting as an Excel workbook. Then reading that back into R!

First, put the iris data into a csv file.

```{r}
(iris_tempfile <- tempfile(pattern = "iris-", fileext = ".csv"))
write.csv(iris, iris_tempfile, row.names = FALSE)
```

Use `googledrive::drive_upload()` to upload the csv and simultaneously convert to a Sheet.

```{r}
(iris_ss <- drive_upload(iris_tempfile, type = "spreadsheet"))

## visit the new Sheet in the browser, in an interactive session!
drive_browse(iris_ss)
```

Read data from the private Sheet into R.
```{r}
read_sheet(iris_ss, range = "B1:D6")
```

Download the Sheet as an Excel workbook and read it back in via `readxl::read_excel()`.

```{r}
(iris_xlsxfile <- sub("[.]csv", ".xlsx", iris_tempfile))
drive_download(iris_ss, path = iris_xlsxfile, overwrite = TRUE)
readxl::read_excel(iris_xlsxfile)
```

Clean up.

```{r}
file.remove(iris_tempfile, iris_xlsxfile)
```

## Get Sheet metadata or detailed cell data

`sheets_get()` exposes Sheet metadata. It has a nice print method, but there's much more info in the object itself.

```{r}
(mini_gap_meta <- sheets_get(sheets_example("mini-gap")))

str(mini_gap_meta, max.level = 1)

mini_gap_meta$sheets
```

`sheets_cells()` returns a data frame with one row per cell and it gives access to raw cell data sent by the Sheets API.

```{r}
(df <- sheets_cells(sheets_example("deaths"), range = "E5:E7"))
df$cell[[3]]
```

`spread_sheet()` converts data in the "one row per cell" form into the data frame you get from `read_sheet()`, which involves reshaping and column typing.

```{r}
df %>% spread_sheet(col_types = "D")
## is same as ...
read_sheet(sheets_example("deaths"), range = "E5:E7", col_types ="D")
```

## What's yet to come?

Writing to Sheets

## Context

googlesheets4 draws on and complements / emulates other packages in the tidyverse:

  * [googledrive](http://googledrive.tidyverse.org) already provides a fully-featured interface to the Google Drive API. Any "whole file" operations can already be accomplished *today* with googledrive: upload or download or update a spreadsheet, copy, rename, move, change permission, delete, etc. googledrive already supports OAuth2 and Team Drives.
  * [readxl](http://readxl.tidyverse.org) is the tidyverse package for reading Excel files (xls or xlsx) into an R data frame. googlesheets4 takes cues from parts of the readxl interface, especially around specifying which cells to read.
  * [readr](http://readr.tidyverse.org) is the tidyverse package for reading delimited files (e.g., csv or tsv) into an R data frame. googlesheets4 takes cues from readr with respect to column type specification.

*Please note that this project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).
By participating in this project you agree to abide by its terms.*
