---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/tidyverse/googlesheets4.svg?branch=master)](https://travis-ci.org/tidyverse/googlesheets4)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/tidyverse/googlesheets4?branch=master&svg=true)](https://ci.appveyor.com/project/tidyverse/googlesheets4)
[![Coverage status](https://codecov.io/gh/tidyverse/googlesheets4/branch/master/graph/badge.svg)](https://codecov.io/github/tidyverse/googlesheets4?branch=master)

# googlesheets4

The goal of googlesheets4 is to ...

## Installation

You can install the released version of googlesheets4 from [CRAN](https://CRAN.R-project.org) with:

``` r
## NO, NO YOU CANNOT
## install.packages("googlesheets4")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tidyverse/googlesheets4")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```

Please note that this project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.
