---
title: "Get started with googlesheets4"
---

```{r setup, include = FALSE}
can_decrypt <- gargle:::secret_can_decrypt("googlesheets4")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = can_decrypt,
  eval = can_decrypt
)
options(tibble.print_min = 5L, tibble.print_max = 5L)
```

```{r eval = !can_decrypt, echo = FALSE, comment = NA}
message("No token available. Code chunks will not be evaluated.")
```

```{r index-auth, include = FALSE}
googlesheets4:::sheets_auth_docs()
```

```{r}
library(googlesheets4)
```

This article takes a quick tour of the main features of googlesheets4. 
Remember to see [the articles](https://googlesheets4.tidyverse.org/articles/index.html) for more detailed treatment of all these topics and more.

## `read_sheet()`, a.k.a. `sheets_read()`

`read_sheet()` is the main "read" function and should evoke `readr::read_csv()` and `readxl::read_excel()`. It's an alias for `sheets_read()`, because most functions in googlesheets4 actually start with `sheets_`. googlesheets4 is pipe-friendly (and reexports `%>%`), but works just fine without the pipe.

`read_sheet()` is designed to "just work", for most purposes, most of the time. It can read straight from a Sheets browser URL:

```{r}
sheets_read("https://docs.google.com/spreadsheets/d/1U6Cf_qEOhiR9AZqTqS3mbMF3zt2db48ZP5v3rkrAEJY/edit#gid=780868077")
```

However, these URLs are not pleasant to work with. More often, you will want to identify a Sheet by its ID:

```{r}
sheets_read("1U6Cf_qEOhiR9AZqTqS3mbMF3zt2db48ZP5v3rkrAEJY")
```

or by its name, which requires an assist from the googledrive package ([googledrive.tidyverse.org](https://googledrive.tidyverse.org)):

<!-- remove the 'message = i' later -->

```{r message=4}
library(googledrive)

drive_get("gapminder") %>% 
  sheets_read()
```

Note that the name-based approach above will only work if **you** have access to a Sheet named "gapminder". Sheet names cannot be used as absolute identifiers; only a Sheet ID can play that role.

For more Sheet identification concepts and strategies, see the article [Find and Identify Sheets](https://googlesheets4.tidyverse.org/articles/articles/find-identify-sheets.html). See the article [Read Sheets](https://googlesheets4.tidyverse.org/articles/articles/find-identify-sheets.html) for more about reading from a specific (work)sheet or ranges, setting column type, and getting low-level cell data.

## Example Sheets and `sheets_browse()`

We've made a few Sheets available to "anyone with a link", for use in examples and docs. Two helper functions make it easy to get your hands on these file IDs.

  * `sheets_examples()` lists all the example Sheets and it can also filter by
    matching names to a regular expression.
  * `sheets_example()` requires a regular expression and returns exactly 1 Sheet
    ID (or throws an error).
    
```{r}
sheets_example("chicken-sheet") %>% 
  sheets_read()
```

If you'd like to see a Sheet in the browser, including our example Sheets, use `sheets_browse()`:

```{r eval = FALSE}
sheets_example("deaths") %>%
  sheets_browse()
```

## Sheet metadata

`sheets_get()` exposes Sheet metadata, such as details on worksheets and named ranges.

```{r}
ss <- sheets_example("deaths")

sheets_get(ss)

sheets_sheet_properties(ss)

sheet_names(ss)
```

`sheets_sheet_properties()` and `sheet_names()` are two members of a larger family of functions for dealing with the (work)sheets within a (spread)Sheet.

The metadata exposed by `sheets_get()` is also revealed whenever you print an object that is known to be a `sheets_id` (an S3 class we use to make Sheet IDs).

`sheets_get()` is related to `googledrive::drive_get()`. Both functions return metadata about a file on Google Drive, such as its ID and name. However, `sheets_get()` reveals additional metadata that is specific to Drive files that happen to be Sheets, such as info about worksheets and named ranges.

## Writing Sheets

*The writing functions are the most recent additions and are likely to see some refinements re: user interface and which function does what.*

`sheets_write()` writes a data frame into a Sheet. The only required argument is the data.

```{r}
df <- data.frame(x = 1:3, y = letters[1:3])

ss <- sheets_write(df)
ss
```

You'll notice the new (spread)Sheet has a randomly generated name. If that is a problem, use `sheets_create()` instead, which affords more control over various aspects of the new Sheet.

Let's start over: we delete that Sheet and call `sheets_create()`, so we can specify the new Sheet's name.

```{r}
googledrive::drive_rm(ss)

ss <- sheets_create("fluffy-bunny", sheets = df)
ss
```

`sheets_write()` can write to new or existing (work)sheets in this Sheet. Let's write the `chickwts` data to a new sheet in `ss`.

```{r}
sheets_write(chickwts, ss)
ss
```

We can also use `sheets_write()` to replace the data in an existing sheet.

```{r}
sheets_write(data.frame(x = 4:10, letters[4:10]), ss, sheet = "df")
ss
```

`sheet_append()` adds one or more rows to an existing sheet.

```{r}
ss %>% sheet_append(data.frame(x = 11, letters[11]), sheet = "df")
ss
```

`range_write()` makes it possible to write arbitrary data, into an arbitrary range. It has a very different "feel" from `sheets_create()`, `sheets_write()`, and `sheet_append()`, all of which assume we're writing or growing a table of data in a (work)sheet. `range_write()` make fewer assumptions about what it's writing and why.

There is also a family of `sheets_sheet_*()` functions that do pure (work)sheet operations, such as add and delete.

We take one last look at the sheets we created in `ss`, then clean up.

```{r}
sheets_sheet_properties(ss)

googledrive::drive_rm(ss)
```

The article [Write Sheets](https://googlesheets4.tidyverse.org/articles/articles/write-sheets.html) has even more detail.
