---
title: "Example Sheets"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  error = TRUE
)
```

An index of the official example and test Sheets. They are world-readable, so we do `gs4_deauth()`.

```{r setup}
library(googlesheets4)

gs4_deauth()
```

## Example Sheets

```{r include = FALSE}
example_urls <- sapply(gs4_examples(), gs4_browse)
links <- glue::glue("[{names(example_urls)}]({unname(example_urls)})")
dat <- tibble::tibble(
  `name (these are links)` = links,
  id = unclass(gs4_examples())
)
```

```{r echo = FALSE}
knitr::kable(dat)
```

### How to get hold of the example Sheets yourself

`gs4_examples()` returns a named vector of Sheet IDs. It is also an instance of `drive_id`, to make good things happen with googledrive.

```{r}
gs4_examples()

gs4_examples("gap")

gs4_examples("and")
```

`gs4_example()` returns exactly one Sheet ID (or errors). It is an instance of `drive_id`. It is also an instance of `sheets_id`, which means printing will (try to) reveal current metadata about the Sheet.

```{r}
gs4_example("chicken")

gs4_example("gap")
```

Here's a handy snippet to open all the example Sheets in browser tabs.

```{r eval = FALSE}
lapply(gs4_examples(), gs4_browse)
```

## Test Sheets

These are more developer-facing, but it's convenient to do them here too.

```{r include = FALSE}
test_urls <- googlesheets4:::.test_sheets %>% 
  purrr::map(gs4_browse)
links <- glue::glue("[{names(test_urls)}]({unname(test_urls)})")
dat <- tibble::tibble(
  `name (these are links)` = links,
  id = unclass(googlesheets4:::.test_sheets)
)
```

```{r echo = FALSE}
knitr::kable(dat)
```
