---
title: "Fun with googledrive and readxl"
---

This article demonstrates how to use googlesheets4, googledrive, and readxl together. We demonstrate a roundtrip for data that starts and ends in R, but travels in spreadsheet form, via Google Sheets.

```{r setup, include = FALSE}
can_decrypt <- gargle:::secret_can_decrypt("googlesheets4")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = can_decrypt,
  eval = can_decrypt
)
```

```{r eval = !can_decrypt, echo = FALSE, comment = NA}
message("No token available. Code chunks will not be evaluated.")
```

## Attach packages

```{r}
library(googlesheets4)
library(googledrive)
library(readxl)
```

## Auth

```{r include = FALSE}
# happens in .onLoad() when IN_PKGDOWN, but need this for local dev/preview
googlesheets4:::gs4_auth_docs(drive = TRUE) 

# attempt to reduce quota exhaustion problems
if (identical(Sys.getenv("IN_PKGDOWN"), "true")) Sys.sleep(20)
```

As a regular, interactive user, you can just let googlesheets4 prompt you for anything it needs re: auth.

Since this article is compiled noninteractively on a server, we activate a service token here, in a hidden chunk. We are also using a shared token for Sheets and Drive. You can read how to do that in your own work in the article [Using googlesheets4 with googledrive](https://googlesheets4.tidyverse.org/articles/articles/drive-and-sheets.html).

## Create a private Sheet from csv with the Drive API

Put the iris data into a csv file.

```{r}
(iris_tempfile <- tempfile(pattern = "iris-", fileext = ".csv"))
write.csv(iris, iris_tempfile, row.names = FALSE)
```

Use `googledrive::drive_upload()` to upload the csv and simultaneously convert to a Sheet.

```{r}
(iris_ss <- drive_upload(iris_tempfile, type = "spreadsheet"))

# visit the new Sheet in the browser, in an interactive session!
drive_browse(iris_ss)
```

Read data from the private Sheet into R.
```{r}
read_sheet(iris_ss, range = "B1:D6")
```

## Create a local xlsx from a Sheet with the Drive API

Download the Sheet as an Excel workbook.

```{r}
(iris_xlsxfile <- sub("[.]csv", ".xlsx", iris_tempfile))
drive_download(iris_ss, path = iris_xlsxfile, overwrite = TRUE)
```

## Read xlsx with readxl

Read the iris data back in via `readxl::read_excel()`.

```{r}
if (requireNamespace("readxl", quietly = TRUE)) {
  readxl::read_excel(iris_xlsxfile)  
}
```

## Clean up

```{r}
file.remove(iris_tempfile, iris_xlsxfile)
drive_trash(iris_ss)
```
