---
title: "Function and class names"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article explains conventions in googlesheets4 around:

  * Function names
  * Classes

# Prefixes: gs4, sheet, range

Almost all functions in googlesheets4 have one of these prefixes:

  * `gs4_`, referring variously to the googlesheets4 package, v4 of the Google 
    Sheets API, or to operations on one or more (spread)Sheets
  * `sheet_`, referring to an operation on one or more (work)sheets
  * `range_`, referring to an operation on a range of cells

This table summarizes what the `gs4_`, `sheet_`, `range_` mean conceptually and which arguments you can expect to see in the function signature.

| prefix | `ss`  | `sheet` | `range` | scope            |
|--------|-------|---------|---------|------------------|
| gs4_   | yes   | no      | no      | a (spread)Sheet  |
| sheet_ | yes   | yes     | no      | a (work)sheet    |
| range_ | yes   | yes     | yes     | a range of cells |

Take this table with a grain of salt.
There are a few violations of it in function signatures, when justified, but it's true in broad strokes.
And remember that `gs4_` is also used for general, package-level functions.

# Previous use of `sheets_` prefix

When googlesheets4 first appeared on CRAN and up to v0.1.1, it had a nearly universal `sheets_` prefix.
In version 0.2.0, googlesheets4 gained a tremendous amount of writing and editing functionality.
At that time, it became clear that the `sheets_` scheme was an impediment to generating descriptive, predictable function names.
For example, you can delete a (spread)Sheet, a (work)sheet, or a cell range.
Which one of these functions will be named `sheets_delete()`?
How do we name the others?
There is no good answer to this, which is why we adopted the 3 prefix strategy.

Any function that existed in v0.1.1 has been formally deprecated and is addressed by formal documentation mapping old names to new.
Below are larger tables, that also cover functions that only existed in dev versions, but that may have seen use in the community.
As time goes on, this article becomes more of historical note.

The dev version of googlesheets4 was bumped from 0.1.1.9000 to 0.1.1.9100 when the `sheets_` naming scheme was abandoned.
The last commit under the old scheme was also tagged as "v0.1.1.9000", which means one can install from that specific state via `devtools::install_github("tidyverse/googlesheets4@v0.1.1.9000")`.

## Auth and the Sheet API v4 surface 

```{r echo = FALSE}
dat <- tibble::tribble(
  ~ "<= v0.1.1.9000",              ~ ">= v0.1.1.9100",
  "`sheets_api_key()` (*)",        "`gs4_api_key()`",
  "`sheets_auth()` (*)",           "`gs4_auth()`",
  "`sheets_auth_configure()` (*)", "`gs4_auth_configure()`",
  "`sheets_deauth()` (*)",         "`gs4_deauth()`",
  "`sheets_endpoints()` (*)",      "`gs4_endpoints()`",
  "`sheets_has_token()` (*)",      "`gs4_has_token()`",
  "`sheets_oauth_app()` (*)",      "`gs4_oauth_app()`",
  "`sheets_token()` (*)",          "`gs4_token()`",
  "`sheets_user()` (*)",           "`gs4_user()`"
)
knitr::kable(dat)
```

(*) indicates functions in the CRAN versions <= v0.1.1.

## (Spread)sheet scope

```{r echo = FALSE}
dat <- tibble::tribble(
  ~ "<= v0.1.1.9000",       ~ ">= v0.1.1.9100",
  "`sheets_browse()` (*)",   "`gs4_browse()`",
  "`sheets_create()`",       "`gs4_create()`",
  "`sheets_find()` (*)",     "`gs4_find()`",
  "`sheets_fodder()`",       "`gs4_fodder()`",
  "`sheets_formula()`",      "`gs4_formula()`",
  "`sheets_example()` (*)",  "`gs4_example()`",
  "`sheets_examples()` (*)", "`gs4_examples()`",
  "`sheets_get()` (*)",      "`gs4_get()`",
  "`sheets_random()`",       "`gs4_random()`"
)
knitr::kable(dat)
```

(*) indicates functions in the CRAN versions <= v0.1.1.

## (Work)sheet scope

```{r echo = FALSE}
dat <- tibble::tribble(
  ~ "<= v0.1.1.9000",            ~ ">= v0.1.1.9100",
  "`sheets_append()`",           "`sheet_append()`",          
  "`sheets_sheet_add()`",        "`sheet_add()`",
  "`sheets_sheet_copy()`",       "`sheet_copy()`",
  "`sheets_sheet_delete()`",     "`sheet_delete()`",
  "`sheets_sheet_names()`",      "`sheet_names()`",
  "`sheets_sheet_properties()`", "`sheet_properties()`",
  "`sheets_sheet_relocate()`",   "`sheet_relocate()`",
  "`sheets_sheet_rename()`",     "`sheet_rename()`",
  "`sheets_sheet_resize()`",     "`sheet_resize()`",
  "`sheets_sheets()` (*)",       "`sheet_names()`",
  "`sheets_write()`",            "`sheet_write()`"
)
knitr::kable(dat)
```

(*) indicates functions in the CRAN versions <= v0.1.1.

## Range scope

```{r echo = FALSE}
dat <- tibble::tribble(
  ~ "<= v0.1.1.9000",            ~ ">= v0.1.1.9100",
  "`sheets_auto_resize_dims()`", "`range_autofit()`",          
  "`sheets_cells()` (*)",        "`range_read_cells()`",
  "`sheets_clear()`",            "`range_clear()`",
  "`sheets_edit()`",             "`range_write()`",
  "`sheets_flood()`",            "`range_flood()`",
  "`sheets_read()` (*)",         "`range_read()`",
  "`sheets_speedread()`",        "`range_speedread()`"
)
knitr::kable(dat)
```

(*) indicates functions in the CRAN versions <= v0.1.1.

Note: "range" can mean two things in the Sheets API:

  * An A1-style range, given as a string. Examples: `A1:B3` (a cell range), `Africa` (a worksheet name), `Africa!A1:B3` (a sheet-qualified cell range), `arts_data` (a named range). Some API endpoints require this, believe it or not.
  * A cell rectangle described by start/end row/column, packaged as an instance of a schema, such as `GridRange`. Most API endpoints use this.

Fun fact: The "cell rectangle" type of range is almost a sub-case of the A1-style range, except there are rectangles open on one or more sides that can be described via `GridRange` that cannot be expressed as an A1-range string.
The mostly-developer-facing article [Range specification](https://googlesheets4.tidyverse.org/articles/articles/range-specification.html) documents all this and how I approach it internally.

The `range_` prefix encompasses both types of ranges and each function has to indicate what sorts of `range` it supports, which is determined by the behaviour of the associated API endpoint.

# Classes

Any user facing class that is related to a schema should be named like `googlesheets4_schema_name`, where the schema name is in lower_snake_case.

The internal schema-derived classes should be like `googlesheets4_schema_SchemaName` / `googlesheets4_schema` / `list`. Use the `googlesheets4_schema` prefix and retain the API's UpperCamelCase.
