---
title: "googlesheets4 design notes"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
```

*This is useful to me and not necessarily for outside consumption.*

## Peer group

Packages to consider in design of googlesheets4:

  * googlesheets: let's keep the good and fix the bad
  * readr: reference point for UI, col spec, parsing
  * readxl: reference point for parsing, (work)sheets/tabs, cell ranges

## Col spec

All 3 packages provide a way for user to specify the output type of each column.

readr has the col spec system that we are most commited to long-term. `col_types` accepts a string using a single letter shorthand (e.g. `"c"` for character), as well as a longer form more conducive to passing ancillary info, such as datetime format. If no column type is given, it is guessed.

googlesheets exposes the readr col spec API. All `gs_read*()` functions literally call either `read_csv()` or `type_convert()`. This has downsides, primarily the fact that, by default, googlesheets reads and parses *formatted* cell contents. The good news is that googlesheets did not introduce its own way to describe col spec.

```{r, cache = TRUE}
library(googlesheets)
gs_read(gs_gap(), ws = "Europe", n_max = 4, col_types = "cciddd")
```

readxl has its own col spec system, but it is slated to evolve towards readr. `col_types` is a character vector with one entry per column from these options: "skip", "guess", "logical", "numeric", "date", "text" or "list". It lacks some of readr's output types and control: no integers (Excel itself has no concept of integer, but still), no factors, less variety wrt date/time/datetime. The "list" type is unique to readxl and is equally relevant to Google Sheets.

```{r, cache = TRUE}
library(readxl)
(clippy <- read_excel(
  readxl_example("clippy.xlsx"),
  col_types = c("text", "list"))
)
tibble::deframe(clippy)
```

## Parsing

Given a column type, each cell is processed to produce an atom in a vector of the target type. This is a loop over cells (within a loop over columns).

readxl: We infer type for every cell, either in the guessing pass or at parse time. Cell type inference is based on Excel's declared cell type, the cell data, and user inputs `na` and `trim_ws` (e.g., https://github.com/tidyverse/readxl/blob/c02f81c3b9f7c6177ed9d13748377450bff9abe1/src/XlsxCell.h#L56).

The cell type and the column type are both enums:

``` cpp
enum CellType {
  CELL_UNKNOWN,
  CELL_BLANK,
  CELL_LOGICAL,
  CELL_DATE,
  CELL_NUMERIC,
  CELL_TEXT
};

enum ColType {
  COL_UNKNOWN, // implies column type needs to be guessed
  COL_BLANK,   // occurs when col_types = NULL and observe only CELL_BLANKs
  COL_LOGICAL,
  COL_DATE,
  COL_NUMERIC,
  COL_TEXT,
  COL_LIST,    // occurs only as user-specified column type
  COL_SKIP     // occurs only as user-specified column type
};
```

There is a large `switch()` statement that fits a cell of type *x* into a column of type *y*. This is simple for matching types, such as `CELL_TEXT` and `COL_TEXT`. When it's not, we must coerce. The simplest coercions are when `NA` is the only sensible value, such as `CELL_LOGICAL` to `COL_DATE`. Examples of trickier coercions: `CELL_TEXT` to `COL_LOGICAL` (we use R's rules for this) or `CELL_DATE` to `COL_TEXT` (we can't really do this right now).

readr. I don't know it well enough to give same summary. Will do if becomes necessary. The googlesheets4 parsing is more likely to be modelled after readxl anyway, because the info available for each cell is more similar. Presumably there is no notion of "cell type" for readr, since it is always a string.

List of the readr parsers (single character shorthand in parens):

  - `col_skip()` (_) (-)
  - `guess_parser()` (?)
  - `parse_logical()` (l)
  - `parse_integer()` (i)
  - `parse_double()` (d)
  - `parse_number()` (n)
  - `parse_datetime(..., format = "??")` (T)
  - `parse_date(..., format = "??")` (D)
  - `parse_time(..., format = "??")` (t)
  - `parse_factor(..., levels = "??")` *why does this not have shortcode? issue opened*
  - `parse_character()` (c)

## Guessing

readxl guesses a col type like so:

  * Determine cell type = cell enum value.
  * Track cumulative max enum value for the first `guess_max` cells. Bit of hand-waving here because, while it's clear that Boolean < integer < double, datetimes are just special.
  * The max enum value is the guessed col type.

readr: I haven't looked at this. I assume it is also a "max col type" approach.

Observation: It makes sense and is very R-like to take the "maximum" type, i.e. the current pushes everything towards character. Is it worth considering a "majority rules" mentality?

## Google Sheet cell data from Sheets API v4

Two options:

  * [spreadsheets.values collection](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets.values): simplest for reading/writing spreadsheet values. Would be nice to use this, but ...
  * [spreadsheets collection](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets): for reading/writing any aspect of the spreadsheet. I *must* use this, at least for reading. Cell typing and coercion demand that we have the contents in various forms (e.g., as entered, unformatted, and formatted), plus information about the format itself.
  
In everything below, I'm suppressing the fact that it's possible to limit the read to specific cells. Not relevant for picking the endpoint.
  
### Example Sheet

I've built an example Sheet with numeric data, especially datetimes and datetime formulas, to study what's available for a cell.

<https://docs.google.com/spreadsheets/d/1xTUxWGcFLtDIHoYJ1WsjQuLmpUtBf--8Bcu5lQ302SU/edit#gid=0>

### spreadsheets.values.get

*This can be skipped, because we will not use this endpoint. But I needed to take a look. It is also the easiest way to juxtapose: as entered, unformatted, formatted.*

[`spreadsheets.values.get`](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets.values/get) has three query parameters:

  * `majorDimension`: whether row or column increments fastest in the reply. Boring. Not explored.
  * `valueRenderOption`: An enum:
    - `FORMATTED_VALUE` (default): Values calculated & formatted according to the cell's formatting (automatic or explicit) and spreadsheet's locale. You basically get the strings that user sees in the browser. Current googlesheets package works with v3 equivalent of this and processes via readr. 
    - `UNFORMATTED_VALUE`: Values are calculated but not formatted. Except there's a loophole: iff the cell has a date-ish format and `dateTimeRenderOption = "FORMATTED_STRING"`, then those specific cells are handled as if `valueRenderOption = "FORMATTED_VALUE"`. I don't see any clever way to exploit this, but just note it.
    - `FORMULA`: Values are not calculated. You get what was entered. If the cell contains a formula, you get that. For non-formula cells, seems to be equivalent to`UNFORMATTED_VALUE`. Except for the same loophole noted above for cells bearing a date-ish format when `dateTimeRenderOption = "FORMATTED_STRING"`.
  * `dateTimeRenderOption`: An enum:
    - Ignored if `valueRenderOption = "FORMATTED_VALUE"`, which is the default.
    - `SERIAL_NUMBER` (default): Cells formatted as date, time, datetime, or duration are output as doubles in "serial number" format. The whole number portion of the value (left of the decimal) counts the days since December 30th 1899. The fractional portion (right of the decimal) counts the time as a fraction of the day. 1900 is not treated as a leap year, i.e. Sheets does not reproduce the Lotus 1-2-3 bug.
    - `FORMATTED_STRING`: Date-ish cells are returned as formatted strings, even if `valueRenderOption = "UNFORMATTED_VALUE"`.

*hidden chunk here that hits the `spreadsheets.values.get` endpoint and gets response ready for inspection*

```{r, include = FALSE, cache = TRUE}
## This is ugly because ... the package we need is the one we're designing.
library(tidyverse)
library(gargle)

#(x <- googledrive::drive_get("googlesheets4-design-exploration"))
#file_id <- x$id
file_id <- "1xTUxWGcFLtDIHoYJ1WsjQuLmpUtBf--8Bcu5lQ302SU"

get_values <- function(valueRenderOption = "FORMATTED_VALUE",
                       dateTimeRenderOption = "SERIAL_NUMBER",
                       spreadsheetId = "1xTUxWGcFLtDIHoYJ1WsjQuLmpUtBf--8Bcu5lQ302SU",
                       range = "Sheet1") {
  req <- gargle::request_build(
    method = "GET",
    path = "v4/spreadsheets/{spreadsheetId}/values/{range}",
    params = list(
      spreadsheetId = spreadsheetId,
      range = range,
      valueRenderOption = valueRenderOption,
      dateTimeRenderOption = dateTimeRenderOption
    ),
    base_url = "https://sheets.googleapis.com",
    key = gargle_api_key()
  )
  raw_resp <- gargle::request_make(req)
  httr::content(raw_resp)[["values"]]
}

f <- function(x) {
  x %>% 
    transpose %>% 
    simplify_all() %>% 
    set_names(map_chr(., 1)) %>% 
    map(`[`, -1) %>% 
    as_tibble()
}

vro <- c("FORMATTED_VALUE", "UNFORMATTED_VALUE", "FORMULA") %>% set_names()
sv_serial <- vro %>%
  map(get_values, dateTimeRenderOption = "SERIAL_NUMBER") %>%
  map(f)
sv_string <- vro %>%
  map(get_values, dateTimeRenderOption = "FORMATTED_STRING") %>%
  map(f)

val_df <- bind_cols(
  sv_serial[["FORMATTED_VALUE"]][c("as_entered", "format")],
  FORMATTED_VALUE = sv_serial[["FORMATTED_VALUE"]][["x"]],
  UNFORMATTED_VALUE = sv_serial[["UNFORMATTED_VALUE"]][["x"]],
  FORMULA = sv_serial[["FORMULA"]][["x"]],
  FORMATTED_VALUE2 = sv_string[["FORMATTED_VALUE"]][["x"]],
  UNFORMATTED_VALUE2 = sv_string[["UNFORMATTED_VALUE"]][["x"]],
  FORMULA2 = sv_string[["FORMULA"]][["x"]]
)
```

Side-by-side comparison of `FORMATTED_VALUE` vs. `UNFORMATTED_VALUE` vs. `FORMULA`. Remember this requires 3 separate API calls, i.e. you can't get all 3 of these from `spreadsheet.values` at the same time.

```{r, echo = FALSE}
val_df %>% 
  select(as_entered, format, FORMATTED_VALUE, UNFORMATTED_VALUE, FORMULA) %>%
  print(n = Inf)
```

Here's a look at same, but combined with `dateTimeRenderOption = "FORMATTED_STRING"`. Observations:

  * `FORMATTED_VALUE` is same as above.
  * `UNFORMATTED_VALUE` is same as `FORMATTED_VALUE` for cells bearing a date-ish format.
  * `FORMULA` is same as above for formula cells and numbers entered as numbers. I note `FORMULA` is different in the case of the datetime that was typed in (first row).

```{r, echo = FALSE}
val_df %>%
  select(as_entered, format, FORMATTED_VALUE2, UNFORMATTED_VALUE2, FORMULA2) %>%
  print(n = Inf)
```

```{r include = FALSE, eval = FALSE}
## backing up claims I make in the prose
identical(val_df$FORMATTED_VALUE, val_df$FORMATTED_VALUE2)     ## TRUE
identical(val_df$UNFORMATTED_VALUE, val_df$UNFORMATTED_VALUE2) ## FALSE
identical(val_df$FORMULA, val_df$FORMULA2)                     ## FALSE

## these should all be cells bearing a date-ish format
val_df %>% 
  filter(UNFORMATTED_VALUE != UNFORMATTED_VALUE2) %>% 
  select(-FORMATTED_VALUE, -FORMATTED_VALUE2, -FORMULA, -FORMULA2)

## there should be no cells bearing a date-ish format
val_df %>% 
  filter(UNFORMATTED_VALUE == UNFORMATTED_VALUE2) %>% 
  select(-FORMATTED_VALUE, -FORMATTED_VALUE2, -FORMULA, -FORMULA2)

## should be non-formula, i.e. literal, datetimes
val_df %>% 
  select(as_entered, format, FORMULA, FORMULA2) %>% 
  filter(FORMULA != FORMULA2)

## everything but literal datetimes
val_df %>% 
  select(as_entered, format, FORMULA, FORMULA2) %>% 
  filter(FORMULA == FORMULA2)
```

### spreadsheets.get

[`spreadsheets.get`](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets/get) can return cell data:

  * `includeGridData = "TRUE"`: Put this in the query to request cell data. You will get *everything*.
  * Field mask. More granular way to specify desired fields in the query. We'll use this in the long-run, for fine control (see end of this article). I've used `includeGridData` here to see full detail (some of which I am still suppressing).
  
*hidden chunk here that hits the `spreadsheets.get` endpoint and gets response ready for inspection*
  
```{r include = FALSE, cache = TRUE}
## This is ugly because ... the package we need is the one we're designing.
library(tidyverse)
library(gargle)

#(x <- googledrive::drive_get("googlesheets4-design-exploration"))
#file_id <- x$id
file_id <- "1xTUxWGcFLtDIHoYJ1WsjQuLmpUtBf--8Bcu5lQ302SU"

req <- gargle::request_build(
  method = "GET",
  path = "v4/spreadsheets/{spreadsheetId}",
  params = list(
    spreadsheetId = file_id,
    ranges = "Sheet1!A:A",
    includeGridData = "true"
  ),
  base_url = "https://sheets.googleapis.com",
  key = gargle_api_key()
)
raw_resp <- gargle::request_make(req)
resp <- httr::content(raw_resp) %>%
  pluck("sheets", 1, "data", 1, "rowData") %>% 
  map(list("values", 1))

## throw away some components that I have no use for, so I can still show
## semi-realistic raw return values, prior to rectangling
## this drops stuff like cell backgroundColor and fontFamily
purge <- function(x) {
  x[["effectiveFormat"]] <- x[["effectiveFormat"]]["numberFormat"]
  x
}
resp <- map(resp, purge)

df <- resp %>% 
  transpose() %>% 
  as_tibble() %>% 
  mutate(
    formattedValue = flatten_chr(formattedValue),
    userEnteredValue_name = map_chr(userEnteredValue, names),
    userEnteredValue_value = map_chr(userEnteredValue, unlist),
    effectiveValue_name = map_chr(effectiveValue, names),
    effectiveValue_value = map_chr(effectiveValue, unlist),
    effectiveFormat_nf_type = map_chr(
      effectiveFormat,
      c("numberFormat", "type"),
      .default = NA
    ),
    effectiveFormat_nf_pattern = map_chr(
      effectiveFormat,
      c("numberFormat", "pattern"),
      .default = NA
    )
  ) %>% 
  select(-userEnteredValue, -effectiveValue, -effectiveFormat)
```

If you drill down, the data you get for each cell is an instance of [`CellData`](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets#CellData).

Here's (trimmed down) `CellData` on cell A2, a literal, explicitly formatted datetime:

```{r}
str(resp[[2]])
```

Here's (trimmed down) `CellData` on cell A3, an automatically formatted datetime formula:

```{r}
str(resp[[3]])
```

Observations:

  * `formattedValue` is useful for coercing to character, i.e. Google has already done it and, unlike Excel, exposes it. It is the same as what you get from `spreadsheets.values.get` when `valueRenderOption = "FORMATTED_VALUE"`.
  * `userEnteredValue` could be used if we aspire to return formulas. It is basically the same as what you get from `spreadsheets.values.get` when `valueRenderOption = "FORMULA"`.
  * `effectiveValue` is our main source of cell data. It is basically the same as what you get from `spreadsheets.values.get` when `valueRenderOption = "UNFORMATTED_VALUE"`.
  * `effectiveValue` and `effectiveFormat`, together, provide the basis for cell typing.
  * `effectiveFormat` *could* provide a basis for transmitting datetime format from the Sheet into R, if we had a lexicon or a grammar for the formats. [Sheets API docs on 
Date and Number Formats](https://developers.google.com/sheets/api/guides/formats). [TimeFormatStrings](https://github.com/WizardMac/TimeFormatStrings), a C library to read and write time format strings (Excel, Stata, Unicode).
  
Let's look at these fields for all cells.

`userEnteredValue` and `effectiveValue` hold an instance of `ExtendedValue`, which is always exactly one of the following:

  * `numberValue`
  * `stringValue`
  * `boolValue`
  * `formulaValue`
  * `errorValue`, which will be an instance of `ErrorValue`

```{r}
df %>%
  select(formattedValue, starts_with("user")) %>% 
  print(n = Inf)
```

```{r}
df %>%
  select(formattedValue, starts_with("effectiveValue")) %>% 
  print(n = Inf)
```

`effectiveFormat`, if it exists, holds an instance of `CellFormat`. We only care about the sub-component `NumberFormat`. If it exists, `NumberFormat` has a `type` and possibly a `pattern`. Possible values for `type` of `NumberFormat`:

  * `TEXT`
  * `NUMBER`
  * `PERCENT`
  * `CURRENCY`
  * `DATE`
  * `TIME`
  * `DATE_TIME`
  * `SCIENTIFIC`

```{r}
df %>%
  select(formattedValue, starts_with("effectiveFormat")) %>% 
  print(n = Inf)
```

A cell will not have an `effectiveFormat` if it bears the `defaultFormat`, which is a [property of the enclosing spreadsheet](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets#spreadsheetproperties).

A cell with an `effectiveFormat` does not necessarily have a `NumberFormat`, indicated by `NA` above. A cell with a `NumberFormat` does not necessarily have a `pattern`, indicated by `NA` above. That means a default pattern for that type is selected based on the user's locale. *[The docs](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets#NumberFormat) say "user's locale" and in other places are careful to say "spreadsheet's locale". Not sure if this is sloppiness or is significant.*

### spreadsheets.values.get vs. spreadsheets.get

Nailing down my understanding of the two endpoints.

Claim: cell values returned by `spreadsheets.values.get` when `valueRenderOption = "FORMATTED_VALUE"` (default) are same as `formattedValue` from `spreadsheets.get`.

```{r}
tibble(
  ss_values_get = val_df$FORMATTED_VALUE,
  ss_get = df$formattedValue[-1],
  same = ss_values_get == ss_get,
  fla = val_df$FORMULA
  ) %>%
  filter(!str_detect(val_df$FORMULA, "now")) %>% 
  select(-fla) %>% 
  print(n = Inf)
```

TRUE. My use of the `now()` datetime formula in the example spreadsheet is regrettable here, hence those rows are filtered out.

Claim: cell values returned by `spreadsheets.values.get` when `valueRenderOption = "UNFORMATTED_VALUE"` are same as `effectiveValue` from `spreadsheets.get`.

```{r}
tibble(
  ss_values_get = val_df$UNFORMATTED_VALUE,
  ss_get = df$effectiveValue_value[-1],
  same = ss_values_get == ss_get,
  fla = val_df$FORMULA
  ) %>%
  filter(!str_detect(val_df$FORMULA, "now")) %>% 
  select(-fla) %>% 
  print(n = Inf)
```

MEH. I suspect this is TRUE up to the width/precision of the double-to-string conversion, which inexplicably seems to be different across the two endpoints. It's too bad that `spreadsheets.values.get` seems to report *greater* precision for the literal datetime than `spreadsheets.get`. It's hard to understand why the endpoint described as more powerful would have lower precision for any sort of cell.

Claim: cell values returned by `spreadsheets.values.get` when `valueRenderOption = "FORMULA"` are same as `userEnteredValue` from `spreadsheets.get`.

```{r}
tibble(
  ss_values_get = val_df$FORMULA,
  ss_get = df$userEnteredValue_value[-1],
  ss_get_type = df$userEnteredValue_name[-1],
  same = ss_values_get == ss_get
  ) %>%
  print(n = Inf)

```

MEH. As above, seems to be same content, but with differences in width/precision of doubles.

### Field mask and partial response

<https://developers.google.com/sheets/api/guides/concepts#partial_responses>

It's clear I will ultimately use `spreadsheets.get` for Sheet reading, so here I explore how to control the payload via a field mask. This seems pretty close to what I'll need to read rectangular regions. Example with the range A2:B3.

```{r, cache = TRUE}
library(tidyverse)
library(gargle)

#(x <- googledrive::drive_get("googlesheets4-design-exploration"))
#file_id <- x$id
file_id <- "1xTUxWGcFLtDIHoYJ1WsjQuLmpUtBf--8Bcu5lQ302SU"

req <- gargle::request_build(
  method = "GET",
  path = "v4/spreadsheets/{spreadsheetId}",
  params = list(
    spreadsheetId = file_id,
    #ranges = "Sheet1!A:A",
    ranges = "Sheet1!A2:B3",
    #includeGridData = TRUE
    fields = "spreadsheetId,properties,spreadsheetUrl,sheets.data(startRow,startColumn),sheets.data.rowData.values(formattedValue,userEnteredValue,effectiveValue,effectiveFormat.numberFormat)"
  ),
  base_url = "https://sheets.googleapis.com",
  key = gargle_api_key()
)
raw_resp <- gargle::request_make(req)
ct <- httr::content(raw_resp)
```

Spreadsheet-level metadata:

```{r}
list(
  spreadsheet_id = "spreadsheetId",
  spreadsheet_url = "spreadsheetUrl",
  title = c("properties", "title"),
  locale = c("properties", "locale"),
  time_zone = c("properties", "timeZone")
) %>%
  map_chr(~ ct[[.x]]) %>% 
  enframe()
```

*Metadata about the sheets is available here, but my current field mask does not request it. I assume we'll populate info about the sheets when registering the Sheet, not while reading rectangles.*

The reply has a component `sheets`, holding instances of `Sheet`, one per sheet given in `ranges` in the request. We have just one because `ranges = "Sheet1!A2:B3"`. We look at the `data` component, which contains an instance of [`GridData`](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets#GridData). 

```{r}
pluck(ct, "sheets", 1, "data", 1) %>% 
  str(max.level = 2)
```

The upper left corner of the rectangle is identified via `startRow` and `startColumn`. Both are zero-based and, apparently, are omitted if the value is zero.

The cell data is found within `rowData`, one object per row and, within that, `values` and, within that, one instance of [`CellData`](https://developers.google.com/sheets/api/reference/rest/v4/spreadsheets#CellData) per cell.

```{r}
resp <- ct %>%
  pluck("sheets", 1, "data", 1, "rowData")
str(resp, max.level = 3)
```

Here it is, post-rectangling.

```{r, echo = FALSE}
library(cellranger)

## modified from https://github.com/rsheets/cellranger/issues/30
unrange <- function(x, major_dimension = c("rows", "columns")) {
  major_dimension <- match.arg(major_dimension)
  limits <- as.cell_limits(x)
  rows <- seq(limits$ul[1], limits$lr[1])
  cols <- seq(limits$ul[2], limits$lr[2])
  rowcol <- switch(
    major_dimension,
    rows = expand.grid(rows = rows, cols = cols, stringsAsFactors = FALSE),
    columns = expand.grid(cols = cols, rows = rows, stringsAsFactors = FALSE)
  )
  cell_addrs <- cell_addr(rowcol[["rows"]], rowcol[["cols"]])
  to_string(cell_addrs, fo = "A1", strict = FALSE)
}

cells <- resp %>% 
  map("values") %>% 
  flatten()

df <- cells %>% 
  transpose(.names = c(
    "formattedValue",
    "userEnteredValue",
    "effectiveValue",
    "effectiveFormat"
  )) %>% 
  as_tibble() %>% 
  mutate(numberFormat = map(effectiveFormat, "numberFormat")) %>% 
  select(-effectiveFormat)
df <- df %>% 
  mutate(enter_type = map_chr(userEnteredValue, names),
         enter_value = map(userEnteredValue, unlist),
         effective_type = map_chr(effectiveValue, names),
         effective_value = map(effectiveValue, unlist),
         formatted_value = flatten_chr(formattedValue),
         format_type = map_chr(numberFormat, "type", .default = NA),
         format_pattern = map_chr(numberFormat, "pattern", .default = NA)) %>% 
  select(-userEnteredValue, -effectiveValue, -formattedValue, -numberFormat) %>% 
  add_column(cell = unrange("A2:B3"), .before = 1)
df
glimpse(df)
```


## Usage and design

Big question: Will googlesheets4 have a notion of a spreadsheet object?

### Peer packages

What do peer packages do?

  * No object. Neither readr nor readxl has a notion of a persistent data source. `read_csv()` (and friends) and `read_excel()` take a path to a file and return data. That is all.
  
    ``` r
    library(readr)
    read_csv(readr_example("mtcars.csv")) %>% head(3)
    
    library(readxl)
    read_excel(readxl_example("datasets.xlsx"), sheet = "mtcars") %>% head(3)
    ```

  * Glorified pointer. googlesheets (the original) has a `googlesheet` object. It holds the file id and other spreadsheet-specific metadata. You pass this object to `gs_read()`.

    ```{r, cache = TRUE}
    library(googlesheets)
    (ss <- gs_gap())
    gs_read(ss, ws = "Europe", n_max = 3)
    ```

  * More glorified pointer. googledrive has a `dribble` object. A key feature is that it's built to hold many files at once, as rows in the dribble. It has no notion of file type. Columns are file id, file title, and a hairy list-column.

  * Persistent, mutable spreadsheet object. The (currently dormant) [rsheets effort](https://github.com/rsheets) spec'd out a *linen object*: an R6 data structure to hold a spreadsheet. Rich Fitzjohn implemented it for Excel spreadsheets, but it was absolutely designed with Google Sheets in mind, too. The spreadsheet object has document-level metadata, holds individual worksheets, and can accomodate cell data and, even, formatting. The idea was to read everything out of the spreadsheet all at once and then work with the linen object for downstream operations. This supports, for example, shared infrastructure for dealing with common spreadsheet aggravations.
  
### More thoughts

A spreadsheet is more like a database with multiple tables than a csv file. Which suggests it's not crazy to think of establishing a connection, then reading from and writing to the different tables within it.

readxl and readr are read-only and the target files tend to be local and static. In contrast, googlesheets4 will be read/write and the target file is more volatile. I feel like this supports the use of a persistent object, but I can't say exactly why.

File id is the only thing that can't change about a Google Sheet. Even if we augment the file id with other metadata, at request time, there's no guarantee that the metadata is still accurate.

  * Storing the file name along with the id just seems humane. It will almost always be correct.
  * It's very nice to have an overview of the worksheets: name, extent. Even readxl will list the worksheet names for you (`readxl::excel_sheets()`).
  * How much do we rely on the most recent metadata versus dutifully update? For example, when asked to read a specific worksheet, do we just try or do we use current or refreshed metadata to first establish that the worksheet actually exists?

These are related, but distinct, and shouldn't be conflated:

  * Registering a spreadsheet as a pointer-with-metadata.
    - Example: current googlesheet object.
  * Capturing an exhaustive snapshot of a spreadsheet or a worksheet as an R object that is intelligible and amenable to further manipulation. The linen object is an example. A lower-tech example is the tibble returned by `googlesheets::gs_read_cellfeed()`. It has one row per cell and columns for, e.g.,  cell reference, various forms of cell contents, and cell formatting.

googlesheets makes data immediately available in two forms:

  * A data frame with same "shape" as the Sheet. Defaults to formatted cell contents, but `gs_read(..., literal = FALSE)` requests unformatted contents.
  * A data frame with one row per cell via `gs_read_cellfeed()`, with cell location data and cell contents in three forms ("literal value" = formatted, "input value" = (sort of) what was typed, "numeric value" = actual numeric value).

In fact, in general `gs_read()` is a call to `gs_read_cellfeed()` followed by a call to `gs_reshape_cellfeed()` (which reshapes and enforces col types). This seems like a good design, i.e. to expose the "one row per cell" form, but offer a nice wrapper that does the reshaping.
