---
title: "How to create a googlesheets4 reprex"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

"reprex" is short for "reproducible example". A reprex is very helpful when reporting a bug or requesting a new feature. The [reprex package](https://reprex.tidyverse.org) helps with some of the fiddly mechanics of making a self-contained, well-formatted reprex.

You might worry about some awkward things when using reprex with googlesheets4:

  * What about auth? What if you need to reprex with a private Sheet?
  * How do you make a Sheet readable by everyone or specific individuals, like
    a package maintainer?

Here we show various ways this can work, if rough order of preference.

## Use a public example Sheet

If you can make your point with one of the example Sheets, exposed via `sheets_examples()` and `sheets_example()`, do so! You can call `sheets_deauth()` explicitly in your reprex to shut down any attempt to get a token.

```{r eval = FALSE}
library(googlesheets4)

sheets_deauth()

# your code goes here ... this is just an example
sheets_example("mini-gap") %>% 
  sheets_get()
```

## Create a Sheet and make it world-readable

If you can create an example Sheet that makes your point and make it readable by "anyone with a link", do so! You can call `sheets_deauth()` explicitly in your reprex to shut down any attempt to get a token.

How do you make a Sheet world-readable? Do this setup once! It should not be part of your reprex. Two options:

  * In the browser: *Share* (big green button) > *Advanced* > *Who has access ... Change* > *Link sharing: On - Anyone with the link*
  * Using googledrive:
  
    ```{r eval = FALSE}
    library(googledrive)
    
    x <- drive_get("YOUR_SHEET_NAME")
    
    drive_share(x, role = "reader", type = "anyone")
    ```

Now make a reprex just like we do with the official example Sheets:  

```{r eval = FALSE}
library(googlesheets4)

sheets_deauth()

# your code goes here ... this is just an example
sheets_get("YOUR_SHEET_ID_OR_URL_GOES_HERE")
```

### Grant access to specific user(s)

If you can't create a world-readable example Sheet, perhaps you can still share one with specific individuals, such as a package maintainer. How to share with specific user(s):

  * In the browser: *Share* (big green button) > *People* > *Enter names or email addresses*.
  * Using googledrive:
  
    ```{r eval = FALSE}
    library(googledrive)
    
    x <- drive_get("YOUR_SHEET_NAME")
    
    drive_share(
      x, 
      role = "reader", type = "user",
      email_address = "jane_package_maintainer@example.org"
    )
    ```

See the next section for advice on your reprex code.

## Use a private Sheet

`reprex::reprex()` eventually runs your code in a fresh, non-interactive R session. You won't be there to do anything about auth, like select the right identity or approve the use of a cached token.

This general situation is documented in the gargle vignette [Non-interactive auth](https://gargle.r-lib.org/articles/non-interactive-auth.html). But here's the short version:

  * Develop a code snippet that works interactively for you, in a clean R
    session.
  * Note who you are logged in as. `sheets_user()` reveals this if you're not
    sure.
  * Insert an explicit call to `sheets_auth(email = "SOMEONE@example.org)`
    into your code (see below).
  * Use `reprex::reprex()`.
  
Here's an example of a snippet suitable for `reprex::reprex()`, assuming the user has successfully run it once interactively, so there's a cached token for "jane_doe@example.com".

```{r eval = FALSE}
library(googlesheets4)

sheets_auth(email = "jane_doe@example.com")
ssid <- "some_very_long_string_of_letters_and_digits"

# your googlesheets4 code goes below here
sheets_get(ssid)
```

If you're reluctant to reveal your email address and/or the spreadsheet id, you can use special comments to create a hidden chunk and a visible body chunk. If the Sheet is private and no one else will be able to access it anyway, this is still a good option to show exactly what you're seeing locally.

```{r eval = FALSE}
library(googlesheets4)

#+ include = FALSE
# code here is executed but the results won't appear in rendered reprex
sheets_auth(email = "jane_doe@example.com")
ssid <- "some_very_long_string_of_letters_and_digits"

#+ include = TRUE
# your googlesheets4 code goes below here
sheets_get(ssid)
```

