---
title: "Messages and errors in googlesheets4"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup}
library(googlesheets4)
```

*In a hidden chunk here, I "export" the internal helpers covered below.*

```{r include = FALSE}
gs4_bullets <- googlesheets4:::gs4_bullets

gs4_abort <- googlesheets4:::gs4_abort
abort_unsupported_conversion <- googlesheets4:::abort_unsupported_conversion
```

## User-facing messages

Everything should be emitted by helpers in `utils-ui.R`: specifically, `gs4_bullets()` and, for errors, `gs4_abort()`.
These helpers are wrappers around `cli::cli_bullets()` and `cli::cli_alert()`, respectively.

```{r}
gs4_bullets(c(
        "noindent",
  " " = "indent",
  "*" = "bullet",
  ">" = "arrow",
  "v" = "Doing good stuff for YOU the user",
  "x" = "Nope nope nope",
  "!" = "You might want to know about this",
  "i" = "The more you know!"
))
```

The helpers encourage consistent styling and make it possible to selectively silence messages coming from googlesheets4.
The googlesheets4 message helpers:

-   Use the [cli package](https://cli.r-lib.org/index.html) to get interpolation, inline markup, and pluralization.
-   Use googlesheets4's custom theme, which has styles for Sheet name, worksheet name, and range.
-   Eventually route through `rlang::inform()`, which is important because `inform()` prints to standard output in interactive sessions. This means that informational messages won't have the same "look" as errors and can generally be more stylish, at least in IDEs like RStudio.
-   Are under the control of the `"googlesheets4_quiet"` option. If it's unset, the default is to show messages (unless we're testing, i.e. the environment variable `TESTTHAT` is `"true"`). `googlesheets4_quiet = TRUE` will suppress messages. There are withr-style convenience helpers: `local_gs4_quiet()` and `with_gs4_quiet()`.

### Inline styling

How we use the inline classes:

-   `.s_sheet` for the name of Google Sheet (custom)
-   `.w_sheet` for the name of a worksheet (custom)
-   `.range` for an A1-style range or named range (custom)
-   `.code` for a column in a data frame and for reserved words, such as `NULL`, `TRUE`, and `NA`
-   `.arg`, `.fun`, `.path`, `.cls`, `.url`. `.email` for their usual purpose

*These may not demo well via pkgdown, but the interactive experience is nice.*

```{r}
nm <- "name-of-a-Google-Sheet"
gs4_bullets(c(v = "Creating new Sheet: {.s_sheet {nm}}"))

nm <- "name-of-a-worksheet"
gs4_bullets(c(v = "Protecting cells on sheet: {.w_sheet {nm}}"))

rg <- "A3:B20"
gs4_bullets(c(v = "Writing to the range: {.range {rg}}"))
```

Above, I don't include a period (`.`) at the end of a message with the form: `Description of thing: THING`.
I view it as a special case of a bullet list of simple items, which also don't get periods.

```{r}
gs4_bullets(c(
  "We're going to list some things:",
  bulletize(gargle_map_cli(month.abb[1:4]))
))
```

Other messages that are complete sentence, or at least aspire to be, **do** get a period.

```{r}
gs4_bullets("Doing the stuff you asked me to do.")

gs4_bullets(c(
  "You probably need to do one of these things:",
  "*" = "Call {.fun some_function}.",
  "*" = "Provide more specific input via {.arg some_argument}."
))
```

Most relevant cli docs:

-   [CLI inline markup](https://cli.r-lib.org/reference/inline-markup.html)
-   [Building a Semantic CLI](https://cli.r-lib.org/articles/semantic-cli.html)

### Pluralization

[cli's pluralization](https://cli.r-lib.org/articles/pluralization.html) is awesome!

```{r}
nm <- "name-of-a-Google-Sheet"
n_new <- 1
gs4_bullets(c(v = "Adding {n_new} sheet{?s} to {.s_sheet {nm}}"))

n_new <- 3
gs4_bullets(c(v = "Adding {n_new} sheet{?s} to {.s_sheet {nm}}"))
```

### Collapsing

[Collapsing lists of things](https://cli.r-lib.org/articles/semantic-cli.html#inline-lists-of-items) is great!
Also more pluralization.

```{r}
new_sheet_names <- c("apple", "banana", "cherry")
gs4_bullets(c("New sheet{?s}: {.w_sheet {new_sheet_names}}"))

new_sheet_names <- "kumquat"
gs4_bullets(c("New sheet{?s}: {.w_sheet {new_sheet_names}}"))
```

### Tricky stuff

If you want to see some tricky examples of building up a message from parts, look here:

-   `range_speedread()`

## Errors

Use `gs4_abort()` instead of `rlang::abort()` or `stop()`.
So far, I'm not really using `...` to put data in the condition, but I could start when/if there's a reason to.

`abort_unsupported_conversion()` is a wrapper around `gs4_abort()`.

```{r}
x <- structure(1, class = c("a", "b", "c"))
abort_unsupported_conversion(x, to = "SheetThing")
```

`abort_unsupported_conversion()` exists to standardize a recurring type of error message, usually encountered during development, not by end-users.
I use it a lot in the default method of an `as_{to}()` generic.

There's not much to add here re: errors, now that `cli::cli_abort()` exists.
Error messages are formed just like informational messages now.
