---
title: "Messages and errors in googlesheets4"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup}
library(googlesheets4)
```

*In a hidden chunk here, I "export" the internal helpers covered below.*

```{r include = FALSE}
gs4_success <- googlesheets4:::gs4_success
gs4_info <- googlesheets4:::gs4_info
gs4_warning <- googlesheets4:::gs4_warning
gs4_danger <- googlesheets4:::gs4_danger

gs4_abort <- googlesheets4:::gs4_abort
abort_unsupported_conversion <- googlesheets4:::abort_unsupported_conversion

dq <- googlesheets4:::dq
sq <- googlesheets4:::sq
bt <- googlesheets4:::bt
```

## User-facing messages

Everything should be emitted by helpers in `utils-ui.R`, such as `gs4_success()` or `gs4_info()`.
These are all wrappers around `cli::cli_alert()` and friends.

```{r}
gs4_success("Doing good stuff")
gs4_info("The more you know!")
gs4_warning("You might want to know about this")
gs4_danger("Something quite bad")
```

The helpers encourage consistent styling and makes it possible to selectively silence messages coming from googlesheets4.
The googlesheets4 message helpers:

-   Use the [cli package](https://cli.r-lib.org/index.html) to get interpolation, inline markup, and pluralization
-   Eventually route through `rlang::inform()`, which is important because `inform()` prints to standard output in interactive sessions. This means that informational messages won't have the same "look" as errors and can generally be more stylish, at least in IDEs like RStudio.
-   Are under the control of the `GOOGLESHEETS4_QUIET` environment variable. If it's unset, the default is to show messages (unless we're testing, i.e. the environment variable `TESTTHAT` is `"true"`). `GOOGLESHEETS4_QUIET=true` will suppress messages. The are convenience helpers `local_gs4_quiet()` and `with_gs4_quiet()`.

### Inline styling

How we use the inline classes:

-   `.file` for the name of Google Sheet
-   `.field` for the name of a worksheet
-   `.field` for an A1-style range or named range
-   `.url` and `.email` for a URL or email address

*These may not demo well via pkgdown, but the interactive experience is nice.*

```{r}
nm <- "name-of-a-Google-Sheet"
gs4_success("Creating new Sheet: {.file {nm}}")

nm <- "name-of-a-worksheet"
gs4_success("Protecting cells on sheet: {.field {nm}}")

rg <- "A3:B20"
gs4_success("Writing to the range {.field {rg}}")
```

Most relevant cli docs:

-   [CLI inline markup](https://cli.r-lib.org/reference/inline-markup.html)
-   [Building a Semantic CLI](https://cli.r-lib.org/articles/semantic-cli.html)

### Line breaks and whitespace

Because cli wants to deal with whitespace and wrapping, for better or for worse, lines breaks and leading whitespace in the source have no effect.
This is different from `glue::glue()`, which we use in error messages (see below), and it's easy to get confused.

```{r}
things <- "rows"
rg <- "A3:B20"
gs4_success(
  "Resizing one
or more {things} in
    {.field {rg}}")
```

### Pluralization

[cli's pluralization](https://cli.r-lib.org/articles/pluralization.html) is awesome!

```{r}
nm <- "name-of-a-Google-Sheet"
n_new <- 1
gs4_success("Adding {n_new} sheet{?s} to {.file {nm}}")

n_new <- 3
gs4_success("Adding {n_new} sheet{?s} to {.file {nm}}")
```

### Collapsing

[Collapsing lists of things](https://cli.r-lib.org/articles/semantic-cli.html#inline-lists-of-items) is great!
Also more pluralization.

```{r}
new_sheet_names <- c("apple", "banana", "cherry")
gs4_success("New sheet{?s}: {.field {new_sheet_names}}")

new_sheet_names <- "kumquat"
gs4_success("New sheet{?s}: {.field {new_sheet_names}}")
```

### Tricky stuff

If you want to see some tricky examples of building up a message from parts, look here:

-   `range_speedread()`
-   `sheet_freeze()`

## Errors

Use `gs4_abort()` instead of `rlang::abort()` or `stop()`.
So far, I'm not really using `...` to put data in the condition, but I could start when/if there's a reason to.
Be prepared to get confused about how to style and line break error messages vs. regular messages, which use `glue::glue()` and `cli::cli_alert()`, respectively.

`abort_unsupported_conversion()` is a wrapper around `gs4_abort()`.

```{r}
x <- structure(1, class = c("a", "b", "c"))
abort_unsupported_conversion(x, to = "foofy")
```

`abort_unsupported_conversion()` exists to standardize a recurring type of error message, usually encountered during development, not by end-users.
I use it a lot in the default method of an `as_{to}()` generic.

### Inline styling

We process the `message` of `gs4_abort()` with `glue::glue()`.

Use helpers `dq()`, `sq()`, and `bt()` for inline style (yes, it's clunky compared to cli):

-   `dq()` for double quotes (Sheet names)
-   `sq()` for single quotes (worksheet names, ranges, and most strings, generally)
-   `bt()` for backticks (argument names, functions)

```{r}
nm <- "name-of-a-Google-Sheet"
gs4_abort("Use double quotes around a Google Sheet name: {dq(nm)}")

nm <- "name-of-a-worksheet"
gs4_abort("Use single quotes around a worksheet name: {sq(nm)}")

gs4_abort("In fact, when in doubt, just use single quotes around {sq('stuff')}")

gs4_abort("
  But use backticks when referring to an {bt('argument')} \\
  or {bt('function()')}")
```

### Multiple lines and bullets

Get multiple lines by sending a character vector as `message`.
Think about what `rlang::format_error_bullets()` does and be intentional with naming: the choices are `i`, `x`, and no name.

```{r}
bad_stuff <- c("eew", "yuck", "uh-oh")
gs4_abort(c(
  "This first line is a header, explaining the general situation",
  x = "Sometimes you have to deliver Very Bad News",
  bad_stuff,
  i = "Maybe a call to {bt('magic_function()')} would help?"
))

endpoint <- "sheets.spreadsheets.WTF"
gs4_abort(c("Endpoint not recognized:", x = "{sq(endpoint)}"))
```

### Line breaks and whitespace

Remember each element of `message` is processed with `glue::glue()`, so let's review the line break situation, which is different to the messages made with cli, which is confusing.

This produces two lines:

```{r}
shift <- "SHIFT"
gs4_abort("
  The {bt('shift')} direction must be specified for this {bt('range')}
  It can't be automatically determined")
```

Use `\\` if you want continuation, i.e. you want one line:

```{r}
shift <- "SHIFT"
gs4_abort("
  The {bt('shift')} direction must be specified for this {bt('range')}. \\
  It can't be automatically determined.")

very_very_very_very_very_very_very_very_long_variable_name <- "HA HA"
gs4_abort(c(
  "Imagine {very_very_very_very_very_very_very_very_long_variable_name} \\
   a line that's long in source but short after interpolation", 
  x = "Short bad thing",
  i = "Helpful tip"
))
```

### Pluralization

Use `cli::pluralize()` directly if you need pluralization in error messages.
Search for `cli::qty()` for a few places where the pluralization was a bit trickier.

```{r}
col_names <- c("apple", "banana")
nc <- 1
gs4_abort(c(
  "Length of {bt('col_names')} is not compatible with the data:",
  x = cli::pluralize("Expected {length(col_names)} un-skipped column{?s}"),
  x = cli::pluralize("But data has {nc} column{?s}")
))

m <- 8
sheets_df <- head(iris)
gs4_abort(c(
  cli::pluralize("There {?is/are} {nrow(sheets_df)} sheet{?s}:"),
  x = "Requested sheet number is out-of-bounds: {m}"
))

sheets_df <- head(iris, 1)
gs4_abort(c(
  cli::pluralize("There {?is/are} {nrow(sheets_df)} sheet{?s}:"),
  x = "Requested sheet number is out-of-bounds: {m}"
))
```

### Collapsing

Use `glue::glue_collapse()` ahead of time, if you need to style and collapse.

```{r}
bad_codes <- letters[1:3]
bad_codes <- glue::glue_collapse(sq(bad_codes), sep = ",")
gs4_abort(c(
  "{bt('col_types')} must be a string of readr-style shortcodes:",
  x = "Unrecognized codes: {bad_codes}"
))
```

### Tricky stuff

If you want to see some tricky examples of building up a message from parts, look here:

-   `rep_ctypes()` (conditional pluralization)

## Future thoughts

Would I ever want to use cli in the errors?
There are good things and bad things about that.
Messing around.
This may very well not show up in pkgdown, which is related to what's problematic.

```{r}
rlang::abort(cli::col_blue("Hello ", "world!"))

rlang::abort(paste("... to highlight the", cli::col_red("search term"),
                   "in a block of text\n"))

error <- cli::combine_ansi_styles("red", "bold")
warn <- cli::combine_ansi_styles("magenta", "underline")
note <- cli::style_italic

rlang::abort(error("Error: subscript out of bounds!\n"))
rlang::abort(warn("Warning: shorter argument was recycled.\n"))
rlang::abort(note("Note: no such directory.\n"))
```
