---
title: "Read Sheets"
---

```{r setup, include = FALSE}
can_decrypt <- gargle:::secret_can_decrypt("googlesheets4")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = can_decrypt,
  eval = can_decrypt
)
```

```{r eval = !can_decrypt, echo = FALSE, comment = NA}
message("No token available. Code chunks will not be evaluated.")
```

```{r}
library(googlesheets4)
```

Basic Sheet reading is shown in the [Get started](https://googlesheets4.tidyverse.org/articles/googlesheets4.html) article. Here we show how to target a specific (work)sheet or cell range, how to deal with column types, and how to get detailed cell data.

## Auth

As a regular, interactive user, you can just let googlesheets4 prompt you for anything it needs re: auth.

Since this article is compiled noninteractively on a server, we have arranged for googlesheets4 to use a service account token (not shown).

```{r include = FALSE}
# happens in .onLoad() when IN_PKGDOWN, but need this for local dev/preview
googlesheets4:::sheets_auth_docs(drive = TRUE)  
```

## Specify the range and column types

Here we read from the "mini-gap" and "deaths" example Sheets to show some of the different ways to specify (work)sheet and cell ranges. Note also that `col_types` gives control of column types, similar to how `col_types` works in readr and readxl.

```{r}
read_sheet(sheets_example("mini-gap"), sheet = 2)

read_sheet(sheets_example("mini-gap"), sheet = "Oceania", n_max = 3)

read_sheet(sheets_example("deaths"), skip = 4, n_max = 10)

read_sheet(
  sheets_example("deaths"), range = "other!A5:F15", col_types = "?ci??D"
)
```

If you looked at the "deaths" spreadsheet in the browser (it's [here](https://docs.google.com/spreadsheets/d/1tuYKzSbLukDLe5ymf_ZKdQA8SfOyeMM7rmf6D6NJpxg/edit#gid=1210215306)), you know that it has some of the typical features of real world spreadsheets: the main data rectangle has prose intended for human-consumption before and after it. That's why we have to specify the range when we read from it.

We've designated the data rectangles as [named ranges](https://support.google.com/docs/answer/63175?co=GENIE.Platform%3DDesktop&hl=en), which provides a very slick way to read them -- definitely less brittle and mysterious than approaches like `range = "other!A5:F15"` or `skip = 4, n_max = 10`. A named range can be passed via the `range =` argument:

```{r}
sheets_example("deaths") %>% 
  read_sheet(range = "arts_data")
```

The named ranges, if any exist, are part of the information returned by `sheets_get()`.

## Detailed cell data

`sheets_cells()` returns a data frame with one row per cell and it gives access to raw cell data sent by the Sheets API.

```{r}
(df <- sheets_cells(sheets_example("deaths"), range = "E5:E7"))
df$cell[[3]]
```

Specify `cell_data = "full", discard_empty = FALSE` to get even more data if you, for example, need access to cell formulas or formatting.

`spread_sheet()` `r emo::ji("wink")` converts data in the "one row per cell" form into the data frame you get from `read_sheet()`, which involves reshaping and column typing.

```{r}
df %>% spread_sheet(col_types = "D")
## is same as ...
read_sheet(sheets_example("deaths"), range = "E5:E7", col_types ="D")
```
