---
title: "Google Sheets API v4"
---

```{r setup}
library(rprojroot)
library(jsonlite)
library(tidyverse)
```

## Exploring the Google Sheets API v4 Discovery Document

This is a machine-generated spec that describes what the API understands. Only the human-edited help strings are patched in.

<a href="https://sheets.googleapis.com/$discovery/rest?version=v4">https://sheets.googleapis.com/$discovery/rest?version=v4!</a>

<https://sheets.googleapis.com/$discovery/rest?version=v4>

*above should be `https://sheets.googleapis.com/$discovery/rest?version=v4` but rendering often URLencodes the `$` and breaks the link*

Perhaps I will use this to automatically generate documentation for a low-level interface provided by the new googlesheets. It will be used internally and exposed for users who want direct access.

Inspired by the way the [node-github](https://mikedeboer.github.io/node-github/) automatically generates docs for the [GitHub API](https://developer.github.com/v3/), based on this [JSON file](https://github.com/mikedeboer/node-github/blob/master/lib/routes.json) documenting the underlying API.

```{r load-endpoints, include = FALSE}
dd_fname <- find_package_root_file("data-raw") %>%
  list.files(pattern = "endpoints-list.rds$", full.names = TRUE)
if (length(dd_fname) < 1) {
  message("Cannot find the endpoint data.")
  knitr::knit_exit()
}
elist <- readRDS(dd_fname)
```


```{r results = "asis", echo = FALSE}
qq <- function(.x, .y = "THINGY!", depth = -1) {
  spaces <- c(2, 4, 8) ## I have no idea why depth * 2 doesn't simply work :(
  pre <- if (depth <= 0) "\n## " else str_c(strrep(" ", spaces[depth]), "* ")
  post <- if (depth <= 0) "\n" else ":"
  if (depth >= 0) {
    cat(str_c(pre, .y, post))
  }
  if (is_list(.x)) {
    cat("\n")
    iwalk(.x, qq, depth = depth + 1)
  } else {
    ## endpoints %>% map_chr("path") %>% nchar() == 55
    cat(
      " ",
      str_trunc(as.character(.x), 60 - spaces[depth] - 2 * depth), "\n")
  }
}
qq(elist)
```

